k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-07 paper.Rmd
next round of editing

Oftentimes, partitioning applications in psychology are focused on between-group similarity rather than within-group heterogeneity, even though both goals oftentimes coincide. High between-group similarity is for example desirable when designing experimental conditions using different stimulus sets that should be as similar as possible with regard to response-relevant attributes [@lahl2006]. Such stimulus selection tasks are usually conducted on the basis of attribute data and not on pairwise (dis)similarity ratings. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. When attribute values are available, anticlustering approaches that do not require a dissimilarity matrix[^dissimilarityfromfeatures] can be used to partition the data. K-means is probably the best-known clustering objective that can directly be computed on attribute values. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize between-group similarity with regard to the mean at the cost of similarity in variance. However, when aiming for overall between-group similarity, neglecting all distribution characteristics other than the mean is misguided; similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

Thus, to obtain an adaptation of the $\mathit{SSE}$ whose maximization can minimize differences in variances, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$:

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to obtain groups that have similar variances across the $P$ attributes. As a caveat, however, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group centroid $\overline{\mathbf{x}}^{(k)}$, but the new variables $y_{ij}$ are calculated using the overall centroid $\overline{\mathbf{x}}$ as reference point. To ensure that $\mathit{SSE_{Var}}$ is a valid criterion, we must therefore ensure that group centroids of the original variables are close to the overall data centroid (and therefore to each other). Ensuring that the group means are close to the overall mean is accomplished by maximizing the classical $\mathit{SSE}$, i.e., via standard k-means anticlustering [@spath1986]. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances in the anticlustering process. By maximizing $\mathit{SSE_{kplus}}$ we seek to obtain a partitioning where means as well as variances are similar between groups. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict rather than in alignment (@ferligoj1992direct). As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. First, it is shown that the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ seem to be rather well aligned: There is a high concentration of partitions where both criteria have high values. At the same time, there are many partitions that fulfil only one of the criteria very well, but fail to also fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

```{r, fig.cap = "The relationship between the objectives $\\mathit{SSE}$ and $\\mathit{SSE_{Var}}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the square maximizes $\\mathit{SSE}$ and thus minimizes the difference in means between groups; the partition represented by the triangular maximizes $\\mathit{SSE_{Var}}$. The partition represented by the diamond maximizes $\\mathit{SSE_{kplus}}$ for $w_1 = w_2 = 1$, and satisfies both constituting criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

@@

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, we compute new variables for each original variable and for each distribution moment that should be considered in the anticlustering process. Analogously to the computation of $\mathit{SSE_{Var}}$, we could then define an $\mathit{SSE_{Skew}}$ using $t = 3$ or an $\mathit{SSE_{Kurtosis}}$ using $t = 4$. However, to formalize the computation of the k-plus criterion for an arbitrary number of distribution moments, some new notation has to be provided.

@@

This computation results in a three dimensional tensor $\mathbf{\hat{X}}_{T \times N \times P}$ = $\{\hat{x}_{tij}\}_{T \times N \times P}$ where---in comparison to the original data matrix $\mathbf{X}_{N \times P}$---the additional dimension is given by the $T$ distribution moments that are considered in the anticlustering process. When defining $\hat{x}_{tij}$, the condition $t = 1$ is separately included to keep the unchanged original data as first entry in the "moment dimension" in $\mathbf{\hat{X}}_{T \times N \times P}$. It is used to equalize the mean---the first moment---between groups.

Using $\mathbf{\hat{X}}_{T \times N \times P}$ and a vector of weights $\mathbf{w} = (w_1, \dots, w_T$), the generalized multicriterion for k-plus anticlustering $\mathit{SSE_{kplusGeneralized}}$ can be formalized:

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$ by setting $w_t = 1$ ($t = 1, \dots, T$); this simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two-dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

When defining the rows of $\mathbf{X'}_{N \times \mathit{TP}}$ with the notation $x'_{i \cdot} = (x'_{i1}, \ldots, x'_{i(TP)})$ ($i = 1, \ldots, N$), the criterion $\mathit{SSE_{kplusGeneralized}}$ can be computed as

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$. Here, computing the k-plus criterion is done by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. A first small scale simulation compared different approaches for weight selection for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights was beyond the scope of the current paper, Simulation 1 investigated the relative performance of a selected number weighting schemes. The results of this initial test determined the benchmark implementation of k-plus anticlustering that was used in Simulation 2. Simulation 2 was a large scale comparison of the k-plus objective and other anticlustering objectives striving for between-group similarity: the diversity and the k-means objective.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. In particular, the initial grouping directly determines the size of the anticlusters, because the exchange procedure following the set-up never changes the sizes of the groups. While the data sets in the simulation studies were always partitioned into equal-sized groups, it would also be possible to ensure that the group sizes differ by 1 at most if an even split is not possible; even different-sized groups can be requested if required by the application.

@@

In Simulation 1, the target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Six weighting schemes were compared. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardised augmented data matrix was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favouring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favouring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

To compare the different weighting schemes, `r length(results)` data sets from were generated from a normal distribution ($M = 0$, $\mathit{SD} = 1$). The number of variables $P$ was determined randomly for each data set and varied between 1 and 5; the number of anticlusters $K$ was also determined randomly for each data set and varied between 2 and 4; the sample size $N$ randomly varied between 20 and 100 with the restriction that $N$ had to be divisible by $K$. For each data set, $\mathit{SSE_{kplusGeneralized}}$ was optimized using the local maximum algorithm, once for each of the six competing weighting schemes.

@@

Table 1 illustrates how well each of the six weighting schemes fulfilled each of the four constituting criteria, averaged across the `r length(results)` simulation runs. For the weighted approaches, the criterion that had received the highest weight during the optimization was fulfilled to the highest degree, as should be expected. Interestingly, the unweighted standardization approach had the best average performance across the four criteria. While the degree of advantage by itself does not seem to be remarkable, this pattern was observed in an astonishing `r mean(unstd_best) * 100`% of all simulation runs, while the level of chance would have been `r 1/6 * 100`% (`r format_p(sig_test$p.value)`). Among the competing weighting schemes, the unweighted standardization approach therefore yielded the most promising results. It also yielded satisfying results with regard to maximizing the standard $\mathit{SSE}$, even slightly outperforming the weighting scheme that explicitly favoured $\mathit{SSE}$. Hence, the unweighted standardization approach implicitly gave high priority to minimizing differences in means. This is a pleasant result because, as discussed earlier, for the higher order moment $\mathit{SSE}$s to be valid, the group means have to be close to the overall mean, which is best accomplished if the standard $\mathit{SSE}$ has high priority.

@@

As a secondary purpose of Simulation 2, I investigated the similarity of partitions obtained when optimizing the different anticlustering objectives. In cluster analysis, partitions identified through different approaches tend to overlap considerably, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each simulation run, i.e., for data set and for each $K$.

@@

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions that were obtained via optimization of the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Moreover, the ARI between two anticlustering partitions hardly differed from the ARI between an anticlustering partition and a random partition. This constitutes an interesting result, illustrating that high between-group similarity can be achieved via very different partitions.

@@

This section demonstrates how readers can easily employ k-plus anticlustering using the R package `anticlust` [@papenberg2020]. The package can be installed in the R environment using the `install.packages()` command.

@@

kplus_anticlustering(
  features,
  K = 9,
  variance = TRUE,
  skew = TRUE,
  kurtosis = FALSE,
  moments = NULL,
  method = "local-maximum",
  standardize = TRUE
)
```

The function call takes about 10 seconds on a contemporary personal computer. The function `kplus_anticlustering()` has three boolean arguments to specify whether variance, skewness and kurtosis should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness and kurtosis have to be specifically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `moments` can be used to specify the desired moments as an integer vector. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features---the original and the augmented variables---are standardized before the optimization starts. Thus, by default, the unweighted standardization approach is used. Whether standardization is used can be adjusted using the boolean argument `standardize`.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and high within-group heterogeneity. In this paper, I presented the k-plus criterion for anticlustering, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: means, variances, skewness, kurtosis, and higher order moments. Depending on the application, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only reflects how similar groups are with regard to their means. Interestingly, when using the unweighted sum approach presented here, the k-plus extension does not change the original k-means criterion, because additional criteria can be incorporated by augmenting the input data.

@@

An interesting mixture of cluster analysis and anticlustering is fair clustering [e.g. @chhabra2021; @delBarrio2022]. Fair clustering performs a cluster analysis under the restriction that the distribution of a set of "protected" attributes---such as race, sex or age---should be similar between clusters. Thus, as compared to pure cluster analysis, fair clustering biases the identification of groups with regard to the protected attributes, striving for algorithmic fairness. A relevant difference to anticlustering is that fair clustering is still concerned with identifying "true" structure in the data, while anticlustering divides data sets into artificial subgroups in such a way that between-group similarity / within-group heterogeneity is maximized. However, these methods are logically similar because anticlustering corresponds to fair clustering when only considering the protected attributes. Hence, a fruitful outlook for research might be to investigate if approaches stemming from the literature on fair clustering can be adopted for anticlustering, and vice versa.

>>>>

Oftentimes, partitioning applications in psychology are focused on between-group similarity rather than within-group heterogeneity, even though both goals usually coincide. High between-group similarity is for example desirable when designing experimental conditions using different stimulus sets that should be as similar as possible with regard to response-relevant attributes [@lahl2006]. Such stimulus selection tasks are usually conducted on the basis of attribute data and not on pairwise (dis)similarity ratings. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. When attribute values are available, anticlustering approaches that do not require a dissimilarity matrix[^dissimilarityfromfeatures] can be used to partition the data. K-means is probably the best-known clustering objective that can directly be computed on attribute values. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize between-group similarity with regard to the mean at the cost of similarity in variance. However, when aiming for overall between-group similarity, neglecting all distribution characteristics other than the mean is misguided; similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

Thus, to obtain an adaptation of the $\mathit{SSE}$ whose maximization can minimize differences in variances between groups, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$:

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to obtain groups that have similar variances across the $P$ attributes. As a caveat, however, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group centroid $\overline{\mathbf{x}}^{(k)}$, but the new variables $y_{ij}$ are calculated using the overall centroid $\overline{\mathbf{x}}$ as reference point. To ensure that $\mathit{SSE_{Var}}$ is a valid criterion, we must therefore ensure that the group centroids of the original variables are close to the overall data centroid (and therefore to each other). Ensuring that the group means are close to the overall mean is accomplished by maximizing the classical $\mathit{SSE}$, i.e., via standard k-means anticlustering [@spath1986]. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

@@

By maximizing $\mathit{SSE_{kplus}}$ we seek to obtain a partitioning where means as well as variances are similar between groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria, if the underlying objectives are in conflict rather than in alignment [@ferligoj1992direct]. As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The criteria seem to be rather well aligned: There is a high concentration of partitions where both the $\mathit{SSE}$ and the $\mathit{SSE_{Var}}$ have high values. At the same time, there are many partitions that fulfil only one of the criteria very well, but fail to also fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

```{r, fig.cap = "The relationship between the objectives $\\mathit{SSE}$ and $\\mathit{SSE_{Var}}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the square maximizes $\\mathit{SSE}$ and thus minimizes the difference in means between groups; the partition represented by the triangular maximizes $\\mathit{SSE_{Var}}$. The partition represented by the diamond maximizes $\\mathit{SSE_{kplus}}$ using $w_1 = w_2 = 1$.", fig.height = 5, fig.width = 5}

@@

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, to extend the k-plus criterion to higher order moments, we compute a new variable for each original variable and for each distribution moment that should be considered in the anticlustering process. Analogously to the computation of $\mathit{SSE_{Var}}$, we could then define an $\mathit{SSE_{Skew}}$ using $t = 3$ or an $\mathit{SSE_{Kurtosis}}$ using $t = 4$. However, to formalize the computation of the k-plus criterion for an arbitrary number of distribution moments, some new notation has to be provided.

@@

This computation results in a three dimensional tensor $\mathbf{\hat{X}}_{T \times N \times P}$ = $\{\hat{x}_{tij}\}_{T \times N \times P}$ where---in comparison to the original data matrix $\mathbf{X}_{N \times P}$---the additional dimension is given by the additional distribution moments that are considered in the anticlustering process. When defining $\hat{x}_{tij}$, the condition $t = 1$ is separately included to keep the unchanged original data as first entry of the "moment dimension" in $\mathbf{\hat{X}}_{T \times N \times P}$. It is used to equalize the mean---the first moment---between groups. Using $\mathbf{\hat{X}}_{T \times N \times P}$ and a vector of weights $\mathbf{w} = (w_1, \dots, w_T$), the generalized multicriterion for k-plus anticlustering $\mathit{SSE_{kplusGeneralized}}$ can be formalized:

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$ by setting $w_t = 1$ ($t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$ can be dropped. Instead, the tensor is flattened into a two-dimensional matrix by appending the newly computed k-plus variables $\hat{x}_{tij}$ ($t > 1$) to the original data matrix $\mathbf{X}_{N \times P}$. We obtain the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

When defining the rows of $\mathbf{X'}_{N \times \mathit{TP}}$ with the notation $x'_{i \cdot} = (x'_{i1}, \ldots, x'_{i(TP)})$ ($i = 1, \ldots, N$), the criterion $\mathit{SSE_{kplusGeneralized}}$ can then be computed as

@@

Thus, when setting the weights $w_t = 1$ ($t = 1, \dots, T$), the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ can be expressed in the same form as the standard k-means criterion $\mathit{SSE}$. Thus, it is possible to define the k-plus criterion by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. Simulation 1 compared a selected number weighting schemes for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. The results of this initial test determined the benchmark implementation of k-plus anticlustering that was used as contestant in the follow-up simulation. Simulation 2 was a large scale comparison of the k-plus objective and other anticlustering objectives striving for between-group similarity: the diversity and the k-means objective.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. In particular, the initial grouping determines the size of the anticlusters, because the exchange procedure following the set-up never changes the sizes of the groups. While the data sets in the simulation studies were always partitioned into equal-sized groups, it would also be possible to ensure that the group sizes differ by 1 at most if an even split is not possible; even different-sized groups can be requested if required by the application.

@@

Six weighting schemes were used to optimize $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, thus minimizing differences with regard to differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The first unweighted approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. The second unweighted approach used the unstandardised augmented data matrix. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favouring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favouring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

In the simulation, `r length(results)` data sets from were generated from a normal distribution ($M = 0$, $\mathit{SD} = 1$). Each data set was divided into groups via optimization of $\mathit{SSE_{kplusGeneralized}}$, once using each of the six weighting schemes. The number of variables $P$ was determined randomly for each data set and varied between 1 and 5; the number of groups $K$ was also determined randomly for each data set and varied between 2 and 4; the sample size $N$ randomly varied between 20 and 100 with the restriction that $N$ had to be divisible by $K$.

@@

Table 1 illustrates how well each of the six weighting schemes fulfilled each of the four constituting criteria, averaged across the `r length(results)` simulation runs. For the weighted approaches, the criterion that had received the highest weight during the optimization was fulfilled to the highest degree, as should be expected. The unweighted standardization approach yielded the best average performance across the four criteria. While the degree of advantage by itself does not seem to be remarkable, this pattern was observed in an astonishing `r mean(unstd_best) * 100`% of all simulation runs, while the level of chance would have been `r 1/6 * 100`% (`r format_p(sig_test$p.value)`). Among the competing weighting schemes, the unweighted standardization approach therefore yielded the most promising results. It performed also remarkably strong with regard to the standard $\mathit{SSE}$, even slightly outperforming the weighting scheme that explicitly favoured the $\mathit{SSE}$. Hence, the unweighted standardization approach implicitly gave high priority to minimizing differences in means. This is a pleasant result because, as discussed earlier, for the higher order $\mathit{SSE}$s to be valid, the group means have to be close to the overall mean, which is best accomplished if the standard $\mathit{SSE}$ has high priority.

@@

As a secondary purpose of Simulation 2, I investigated the similarity of partitions obtained when optimizing the different anticlustering objectives. In cluster analysis, partitions identified through different approaches tend to overlap considerably, because similar elements tend to be placed in the same cluster while dissimilar elements tend to be placed in different clusters; this principle holds regardless of the specific clustering method. It is an open question whether the optimization of different anticlustering objectives leads to similar groupings of the elements in anticlusters. To address this question, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each simulation run, i.e., for data set and for each $K$.

@@

To answer the question if different anticlustering partitions are similar or dissimilar, Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, there was hardly any similarity between anticlustering partitions that were obtained via optimization of the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Even the pairwise ARI of the two k-plus criteria was fractional, although the criteria overlapped considerably in their optimization goal. This constitutes an interesting result, illustrating that high between-group similarity can be achieved via very different partitions.

@@

This section demonstrates how readers can easily apply k-plus anticlustering using the R package `anticlust` [@papenberg2020]. The package can be installed in the R environment using the `install.packages()` command.

@@

kplus_anticlustering(
  features,
  K = 9,
  variance = TRUE,
  skew = TRUE,
  kurtosis = FALSE,
  T = NULL,
  method = "local-maximum",
  standardize = TRUE
)
```

The function call takes about 10 seconds on a contemporary personal computer. The function `kplus_anticlustering()` has three boolean arguments to specify whether variance, skewness and kurtosis should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness and kurtosis have to be specifically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `T` can be used to specify how many moments should be used. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features---the original and the augmented variables---are standardized before the optimization starts. Thus, by default, the unweighted standardization approach is used. Whether standardization is used can be adjusted using the boolean argument `standardize`.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and high within-group heterogeneity. In this paper, I presented the k-plus criterion for anticlustering, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: means, variances, skewness, kurtosis, and higher order moments. Depending on the application, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only reflects how similar groups are with regard to their means. Interestingly, when using the unweighted sum approach presented here, the k-plus extension does not even change the original k-means criterion, because additional criteria can be incorporated by augmenting the input data.

@@

An interesting mixture of cluster analysis and anticlustering is fair clustering [e.g. @chhabra2021; @delBarrio2022]. Fair clustering performs a cluster analysis under the restriction that the distribution of a set of "protected" attributes---such as race, sex or age---should be similar between clusters. Thus, as compared to pure cluster analysis, fair clustering biases the identification of groups with regard to the protected attributes, striving for algorithmic fairness. A relevant difference to anticlustering is that fair clustering is still concerned with identifying "true" structure in the data, while anticlustering divides data sets into purely artificial subgroups. However, these methods are logically similar because anticlustering corresponds to fair clustering when only considering the protected attributes. Hence, a fruitful outlook for research might be to investigate if approaches stemming from the literature on fair clustering can be adopted for anticlustering, and vice versa.
